\subsection{Probabilistic Software Analysis} 
Code-level probabilistic analysis is an active area of research with 
techniques based on abstract interpretation~\cite{Monniaux2000,CousotESOP12}, 
static analysis~\cite{adje-etal-vstte13,Rajamani1,Rajamani2} and symbolic
execution~\cite{ISSTA12,PLDI13}.  All these techniques target 
sequential code and they differ in the type of input
distributions they consider, the language features supported, and
the approach used to calculate the number of solutions.

We will build on our previous work
from~\cite{geldenhuys-etal-issta2012,filieri-etal-icse2013,Borges2014PLDI}, that handles numeric constraints and
complex data structures as inputs (the latter is not described here,
for brevity). The goal of the analysis is: (1) to identify the 
symbolic constraints characterizing the inputs that make the 
execution satisfy a given property, and then (2) to quantify the 
probability of satisfying the constraints.
For simplicity, we assume the satisfaction of the target property to be 
characterized by the occurrence of a target event (e.g. successful 
termination or failure), but our work extends to bounded 
LTL~\cite{Zuliani2010} as well. 


The analysis works with a limited budget of symbolic paths, obtained with a bounded symbolic execution of the program. 
 Some of these paths lead to
failure and some of them to success (termination without failure). 
These path conditions are classified in two disjoint sets: 
$\textit{PC}^s=\{\textit{PC}_1^s, 
\textit{PC}_2^s, \ldots , \textit{PC}_m^s\}$ and
$\textit{PC}^f=\{\textit{PC}_1^f, 
\textit{PC}_2^f, \ldots , \textit{PC}_p^f\}$.
The path conditions 
may not cover the full input domain due to inherent incompleteness in the analysis (e.g. due to non-terminating loops or non-exhaustive path exploration) -- these remaining paths are called  {\em grey} paths and are used in~\cite{filieri-etal-icse2013} to quantify the confidence one can put in the bounded symbolic analysis.

\subsubsection{Probabilistic Usage Profiles} 

The constraints generated with symbolic execution are analyzed to
quantify the likelihood of an input to satisfy them, where the inputs
are distributed according to given {\em usage
profiles}~\cite{filieri-etal-icse2013}. A usage profile is a probabilistic
characterization of the software interactions with the external world,
e.g. the users or the physical execution environment.  It assigns to
each valid combination of inputs its probability to occur during
 execution. 
Usage profiles can come from monitoring the usage of
 actual or similar systems or expert and domain knowledge (physical
 phenomena). 
In~\cite{filieri-etal-icse2013} we assumed that the usage profiles
are given but part of the proposed project will investigate techniques for the 
automated inference of compact usage profiles from the usage data 
available and from the analysis of the program itself.

In~\cite{filieri-etal-icse2013}, we also assumed that all the input variables
range over finite discrete domains, whose joining is generically
indicated as $D$. We relaxed this assumption in more recent work~\cite{Borges2014PLDI}. We profile the expected usage for the program through a profile
$\textit{UP}$, which is a set of pairs $\langle
c_i, p_i \rangle$ where $c_i$ is a \emph{usage scenario} defined as a
(constraint representing a) subset of $D$ and $p_i$ ($p_i\geq 0$) is
the probability that a user input belongs to $c_i$. We further
require, for simplicity, $\{c_i\}$ to be a complete partition of $D$,
and thus $\sum_i p_i=1$. Intuitively, $UP$ is the distribution over
the input space. Notice that $c_i$ could contain even a single element
of $D$, allowing for the finest grained specifications of
$\textit{UP}$.

Given the output of symbolic execution, the probability of success can
be defined as the probability of executing the program ($P$) with
an input satisfying any of the successful path conditions, given the
profile $UP$. This definition can be formalized as
$\textit{Pr}^s(P)=\sum_{i} \textit{Pr}(\textit{PC}_i^s \ | \ \textit{UP})$.
An analogous definition is provided for the probability of failure,
$\textit{Pr}^f(P)$. The probability of grey paths is $1-(\textit{Pr}^s(P)+\textit{Pr}^f(P))$ and it
quantifies the ratio of elements
of the input domain for which neither success nor failure have been
revealed for the current analysis. This information is a measure of
the confidence we can put on the probability estimation, under
the current exploration bound.

\subsubsection{Computing Probabilities using Model Counting} 
To compute the probabilities of path conditions, we use a
quantification procedure for the generated constraints.
In~\cite{filieri-etal-icse2013} we used model counting techniques,
i.e. LattE~\cite{deLoera-etal-2012}, to estimate (algorithmically) the exact
number of points of a bounded (possibly very large) discrete domain
that satisfy linear constraints.  In more recent work~\cite{Borges2014PLDI},  
we developed quantification procedures for
the analysis of programs that have mixed integer and floating point
constraints of arbitrary complexity.

To compute the estimated probability of
success (or failure)  we use the fact that $UP$
defines a partition of the input domain and then, from the law of total
probability~\cite{pestman2009}:
$\textit{Pr}(\textit{PC} \ | \ \textit{UP}) =  \sum_{i} \textit{Pr}(\textit{PC} \ | \ c_i) \cdot p_i$.
Furthermore, from the definition of conditional probability~\cite{pestman2009}:
$\textit{Pr}(\textit{PC} |c_i) = \textit{Pr}(\textit{PC} \wedge c_i) / \textit{Pr}(c_i)$. 

To use model-counting techniques for the computation of the
conditional probabilities, let us define for a constraint $c$ the
function $\sharp(c)$ that returns the number of elements of $D$
satisfying $c$. $\sharp(\cdot)$ is always a finite non negative
integer because we assumed $D$ finite and countable. Under this same
assumption, $\textit{Pr}(c)$ is, by definition~\cite{pestman2009},
$\sharp(c) / \sharp(D)$ (where $\sharp(D)$ is the size of the non-empty
input domain).  Thus, one can express the probability of success as:
\setlength\abovedisplayskip{-1pt}
\setlength\belowdisplayskip{-2pt}
\begin{equation*}
\textit{Pr}^s(P)=\sum_{i} \textit{Pr}(\textit{PC}_i^s \ | \ \textit{UP})=
 \sum_{i} \sum_{j} \textit{Pr}(\textit{PC}_i^s \ | \ c_j) \cdot p_j= \sum_{i} \sum_{j} \frac{\sharp(\textit{PC}_i^s \land c_j)}{\sharp(c_j)} \cdot p_j
\end{equation*}
