%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\documentclass[runningheads,a4paper]{llncs}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{color}
\usepackage{amsmath,amssymb,mathtools}
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{centernot}
\usepackage{tikz}
\usetikzlibrary{calc}
\usetikzlibrary{shapes}
\usepackage{pgfplots}
\usepackage{pgfplotstable}


\newcommand\x[1]{\ensuremath{\mathit{#1}}}
\newcommand\lrangle[1]{\ensuremath{\langle#1\rangle}}

\makeatletter
\providecommand{\bigsqcap}{%
  \mathop{%
    \mathpalette\@updown\bigsqcup
  }%
}
\newcommand*{\@updown}[2]{%
  \rotatebox[origin=c]{180}{$\m@th#1#2$}%
}
\makeatother


%\usepackage{url}
%\urldef{\mailsa}\path|{dwyer, mgerrard}@cse.unl.edu|    
%\newcommand{\keywords}[1]{\par\addvspace\baselineskip
%\noindent\keywordname\enspace\ignorespaces#1}

\begin{document}

\newcommand{\mycomment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\ignore}[1]{}

% first the title is needed
\title{Information for GTTSE'2015 Participants}

\author{Matthew B. Dwyer\inst{1}
\and Antonio Filieri\inst{2}\and Jaco Geldenhuys\inst{4}\and Mitchell Gerrard\inst{1}\and\\
Corina Pasareanu\inst{3}\and Willem Visser\inst{4}}
%
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{University of Nebraska -- Lincoln
\and University of Stuttgart
\and NASA Ames Research Center
\and University of Stellenbosch
}
\maketitle

The following paper is a \textbf{DRAFT} of the final paper that we
plan to prepare.  Our goal is to provide an accessible entrance to the
deep and technical literature on probabilistic program analysis.
As a GTTSE participant you are in a position to help us achieve that
goal.

Towards that end we encourage you to send us an email, at
\texttt{dwyer@cse.unl.edu}, with a title beginning ``Feedback on
GTTSE paper'' where you let us know what you think.
In particular, we welcome suggestions for parts of the development
where we might provide a clearer or more explicit explanation of
the ideas.  

We have a number of ideas along these lines, e.g., replacing some
of the examples with ones that are more complex, adding some visualizations
to supplement textual explanations.   We will mix our ideas with yours
and try to come up with a paper that is useful to the broader community.

We'll be happy to thank you for your contributions so do include your
name in the email.

See you in Braga




\end{document}
